---
title: "Newton’s Method"
author: "Morgan de Ferrante"
date: "January 8, 2018"
output: pdf_document
---


##Newton's Method Formula
Newton's method is a root-finding method that uses the derivative to solve for a given function. 

$$x_n = x_{n-1} - \frac{f(x_{n-1})}{f'(x_{n-1})}$$

The method begins with an initial value $x_0$, which needs to be sufficiently close to the true value of the root. Thus iteration of the method begins with the following: 

$$x_1 = x_{0} - \frac{f(x_0)}{f'(x_0)}$$ 

The following newtons_method function generates a list of the iterative values, and a plot of $x_n$ versus $f(x_n)$. The newtons_method function takes in a function object type (f_and_fprime) with both the given function for which we wish to estimate roots and its derivative function, then the initial $x_0$, and the number of desired iterations.

```{r}
library(ggplot2)

newtons_method <- function(f_and_fprime, x0, iter = 5) {
  # applies Newton-Raphson to find x such that f_and_fprime(x)[1] == 0
  # f_and_fprime is a function of x. it returns two values, f(x) and f'(x)
  # x0 is the starting point
  
  # df_points is used to track each update
df_points <- data.frame(x1 = numeric(0), 
                        y1 = numeric(0),
                        x2 = numeric(0),
                        y2 = numeric(0))  
  xnew <- x0
  cat("Starting value is:", xnew, "\n")
  
  # the algorithm which generates each xn value 
  for (i in 1:iter) {
    xold <- xnew
    f_xold <- f_and_fprime(xold)
    xnew <- xold - f_xold[1]/f_xold[2] 
    #f_xold[1] is the value of the original function at xold
    #f_xold[2] is the value of its derivative at xold 
    cat("Next x value:", xnew, "\n")
    
    df_points[2*i - 1,] <- c(xold, 0 , xold, f_xold[1]) # vertical segment 
    df_points[2*i,] <- c(xnew , 0 ,xold ,f_xold[1])   # tangent segment 
  }

  # determines the limits for the graph  
  x_start <- min(df_points$x1, df_points$x2, x0 - .1) # start is the min of these values
  x_end <- max(df_points$x1, df_points$x2, x0 + .1)  # end is the max of these values
  x <- seq(x_start, x_end, length.out = 200)
  fx <- rep(NA, length(x))
  for (i in seq_along(x)) {
    fx[i] <- f_and_fprime(x[i])
  }
  function_data <- data.frame(x, fx)
  
  # uses ggplot to plot the function and the segments for each iteration
  p <- ggplot(function_data, aes(x = x, y = fx)) + 
    geom_line(colour = "blue", size = 1) + 
    geom_segment(aes(x = x1, y = y1, xend = x2, yend = y2), data = df_points) + 
    geom_abline(intercept = 0, slope = 1) + 
    geom_abline(intercept = 0, slope = 0)
  
  print(p)
  return(cat("The estimated root for the function is", xnew)) 
}
```

##Example

This example estimates the zero for the function $f(x) = log(x) - e^{-x}$ using the initial value of $x_0$ = 2. 

```{r, warning=FALSE} 
f_and_fprime <- function(x){
  f <- log(x) - exp(-x)   # f(x)
  fprime <- 1/x + exp(-x)  # f'(x)
  return(c(f, fprime))
} 
newtons_method(f_and_fprime, 2, iter = 8)

```

